---
title: "My story"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(knitr)
library(kableExtra)
```


# About Vasu Dev Puri

I am currently reading for my Masters in Analytics and Management at London Business School and undergoing rigorous training in academic disciplines surrounding Data Analytics, Strategy & Management. My experience, thus far, has provided latitude to appraise businesses through a sector-agnostic approach, coupling the need to identify opportunistic liquid assets using fundamental & thematic analysis.
 
Covid-19, a rather unfortunate sign of our times, offers an unparalleled up scaling potential arising from exposure to market momentum. Current coursework focused on data analytics, coupled with an appreciation of languages such as R & Python, has lent the opportunity to seek upside capture through analytics driven decisions - based on the systemic accretion & application of data.

Prior to joining LBS, I advised SWFs, Asset Managers & PEs in the capacity of a Consultant at the Valuations and Modelling team at PwC Deals, focusing on Infrastructure (Renewables & Toll Roads), Healthcare & Retail verticals. During this time, I worked on several fair value, business restructuring / turnaround , asset optimization, business strategy, purchase price allocations, synthetic credit ratings and business model review engagements.

From an academia standpoint, I’ve cleared the CFA Level 1, and am currently a CFA level 2 candidate. Furthermore, I have completed my Association of Chartered Certified Accountants, was recognized as the youngest member of the Chartered Institute for Securities and Investments and have completed the Corporate Finance qualification (CISI & ICAEW).

As a certified Neuro-Linguistic Programmer, I incorporate the NLP models at my workplace which allow me to build rapport, dissolve conflicts and remove limiting beliefs among team members. Furthermore, NLP has improved my interpersonal and presentation skills that have resulted in effective communication with clients and colleagues.

I have been a part of the Student Council at PwC's Academy since its inception and have served as the Head of Corporate Responsibility from 2018 to 2019. This position has allowed me to influence people in a positive way by instilling a sense of community and purpose in their day-to-day actions and coordinate effectively between multiple stakeholders (NGOs, Governments, and Students). 

A fun fact?

I'm passionate about public speaking, poetry, polo and equestrian. 




