---
title: "Can Trump trump the 2020 US Presdential Elections?"
author: "Vasu Dev Puri"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     
  size="small")   
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(kableExtra)
library(lubridate)
library(infer)
```


# Trump's Approval Margins

In the following section of the report we will analyze data on [fivethirtyeight.com](https://projects.fivethirtyeight.com/trump-approval-ratings) on all polls that track the president's approval

```{r, cache=TRUE}
 
#Import approval polls data
approval_polllist <- read_csv("https://projects.fivethirtyeight.com/trump-approval-data/approval_polllist.csv")

glimpse(approval_polllist)

# Use `lubridate` to fix dates, as they are given as characters.

approval_polllist$modeldate <- mdy(approval_polllist$modeldate)
approval_polllist$startdate <- mdy(approval_polllist$startdate)
approval_polllist$enddate <- mdy(approval_polllist$enddate)

approval_polllist

```



```{r, cache=TRUE}
# Import approval polls data
approval_polllist <- read_csv('https://projects.fivethirtyeight.com/trump-approval-data/approval_polllist.csv')

glimpse(approval_polllist)
?lubridate
# Use `lubridate` to fix dates, as they are given as characters.

# Converting date to number format, grouping by week and year

converteddates <- mdy(approval_polllist$enddate)
week_no <- week(converteddates)
year_no <- year(converteddates)

 

modified_approval_polllist <- approval_polllist  %>% 
  mutate(converteddates, week_no, year_no)

 

# grouping by week and computing net rate and using the plotrix package to compute std.error

library("plotrix")

modified_approval_polllist <-  modified_approval_polllist %>%
    mutate(Net_ratecalculation = (approve-disapprove)/(approve+disapprove)) %>%
    group_by(year_no, week_no) %>%  
  summarise(average_nrc = mean(Net_ratecalculation)*100, 
            standard_error_of_net_rate = std.error(Net_ratecalculation)*100,
            Degrees_of_freedom = length(Net_ratecalculation)-1)
               


#Computing confidence intervals of net rate

 

modified_approval_polllist <- 
  modified_approval_polllist %>% 
  mutate(
    upper_ribbondata = average_nrc + qt(.975, Degrees_of_freedom)*standard_error_of_net_rate,
    lower_ribbondata_1 = average_nrc - qt(.975, Degrees_of_freedom)*standard_error_of_net_rate)

 

#Plotting the graph

 #creating the color palette


custom_color_palette <- c("#FF7F50" ,"#7FFF00", "#00FFFF", "#9400D3")

 
graph_withoutColor <-  ggplot(modified_approval_polllist) +
  geom_line(aes(x=week_no, y=average_nrc , color = factor(year_no)))+
  geom_ribbon(aes(x=week_no, ymin = lower_ribbondata_1, ymax = upper_ribbondata, fill = factor(year_no)), alpha=0.2)+
 # geom_line(aes(x =  , y = 0 , ), color = 'orange')+
  geom_hline(yintercept = 0 , color = 'orange')+
  geom_point(aes(x = week_no, y = average_nrc, color = factor(year_no)))+
  facet_wrap(~year_no)+
  labs(y= "Average Net Approval (%)", x = "Week of the year") +
ggtitle(label = "Estimating Net Approval (approve - dissaprove) for Donald Trump", subtitle = "Weekly average of all polls") +
  scale_y_continuous(breaks = seq(-20, 7.5 , by = 2.5)) +
  scale_x_continuous(breaks = seq(0, 52, by = 13))+
#plotting the graph!+
scale_colour_manual(aesthetics = "custom_color_palette") 
                                                                                                                
graph_withoutColor  +theme_bw() + theme(legend.position = "none")                                                                           
                                                                                                                                                                                                         
```


## Compare Confidence Intervals

From the graph above, it can be inferred that irrespective of the week, Donald's approval rating remains in the negative territory. A discerning sign, considering that the 2020 elections are right around the corner.

Analysis on the confidence bands of week 15 & 34 of 2020 lends new insights (please refer to the plot below), suggesting that the confidence interval is expanding as the election date approaches, suggesting greater variability in the voter's perception of Trump. This can be a function of severe aberrations in voter perception, owing to socio-economic events such as crisis management during the Covid-19 Pandemic, Black Lives Matter movement & President Obama's speech condoning Republican policies.

   

```{r}

#plot CI for trump approval ratings in week 15 and 34
modified_approval_polllist %>%
  filter((week_no == 15 | week_no == 34) & year_no == 2020 ) %>%
  mutate(color  = ifelse(week_no == 15,  'Yes' , 'no'))%>%
  ggplot() + 
  #plottting error bars 
  geom_errorbar(aes(x = week_no , 
                    y = average_nrc ,
                   ymax  = upper_ribbondata,
                   ymin  = lower_ribbondata_1,
                   width = 0.4,
                   color = color,
                   ),size = 2) +
  geom_point(aes(x = week_no , 
                    y = average_nrc ,
                 color = color,
                 fill = color), shape = 21 , size = 4) +
 labs(y= "Average Net Approval (%)", x = "Week of the year") +
ggtitle(label = "Opinions about Trump vary as election approaches", subtitle = "Average of net approval (approve - disapprove) for Trump on selected weeks") +
  scale_colour_manual(aesthetics = "custom_color_palette") +
  theme_bw() +
  theme(legend.position = 'none')
```

## Conclusions

It remains to be seen, how the 2020 elections will unfold. At this point, spurious voter patterns and negative approval margins suggest a rather gloomy outlook. With dwindling voter ratings, uptick in Covid-19  cases and robust candidacy of Democratic president nominee Joe Biden, it remains to be seen what prizes or perils the winter will bring for Donald. 

```{r}
```